Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan

### 2019-11-05 analysis/paper/paper.Rmd
update the method section

Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approach that captures shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes. Landmarks, curves or outlines of objects can be transformed into coordinates by their unique location and orientation with respect to numerical values of coordinate axes [@Adams2004; @Lawing2010; @Slice2007]. According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004]. Landmark approaches apply landmark points to specimen as variables that can be specified on a coordinate system. According to the specific location of each point, objects are translated and scaled into the same orientation and size using Procrustes superimposition [@Lawing2010]. This method can be applied to 3D objects not just limited to 2D that commonly applied to zooarchaeology and

@@

The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction. The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances. Finally, we computed a significance test to assess the equality of coefficients of variation (CVs) among multiple groups of PCs to understand the differences in shape standardisation among context groups. The R package cvequality (Version 0.2.0; Marwick and Krishnamoorthy 2019) in the CRAN repository was used to conduct the significance test for CVs.

>>>>

Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approach that captures shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes. Landmarks, curves or outlines of objects can be transformed into coordinates by their unique location and orientation with respect to numerical values of coordinate axes [@Adams2004; @Lawing2010; @Slice2007]. According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004]. Landmark approaches assign a set of landmark points to specimen that can be specified on a coordinate system. The landmark coordinates are represented by numerical values with respect to defined coordinate axes. Based on the specific position of each point, objects are translated, rotated, and scaled into the same orientation and size using Procrustes superimposition technique to find the best fit between configuration shapes [@Lawing2010]. Landmarks

In archaeology, the landmarks are according to “type two” landmarks of Bookstein (1991), which define them as points located in the maximum of curvature or extreme points in morphology (Bookstein 1991; Zelditch et al. 2004).

were chosed  be applied to 3D objects not just limited to 2D that commonly applied to zooarchaeology and

the analysis of shape coordinates generated by the least-squares superimposition of configurations of landmarks

@@

The harmonic coefficients generated by EFA were analysed by principal component analysis (PCA) to illustrate the diversity of the shape data and identify the major patterns of variation through dimensionality reduction. The principal components (PCs) scores were analysed with a multivariate analysis of variance (MANOVA) to test for significant effects of the groups of occupation context on shape variances. Finally, we computed coefficients of variation statistic (CVs) among multiple groups of PCs with a significance test for the equality of CVs, which enable us to compare CVs in a statistical way (Karl Pearson 1896). The coefficient of variation is a common and widely-used statistical measure of the spread of a set of measurements of a sample. It is defined as the standard deviation divided by the mean in a ratio scale format.

$$c_v = \frac{\sigma}{\mu}$$

By standardizing standard deviation in each data set, the coefficients of variation statistics allows us to directly compare variation in samples measured with different units or means. Although it is a useful method, there is a question that how should we make a decision that the differences among the multiple CVs are significant or not? To answer this question, we need convenient methods for testing the equality of $c_v$ values from different samples, which has been described in many publications and here we proposed a significant test for CVs using the R package cvequality [@Marwick2019] in the CRAN repository. The package cvequality includes two tests, the "Asymptotic test for the equality of coefficients of variation from k populations" (Feltz and Miller 1996) and the "Modified signed-likelihood ratio test (SLRT) for equality of CVs" (Krishnamoorthy and Lee 2014). Feltz and Miller (1996) test is widely cited as an authoritative test for the equality of $c_v$ values that performs better than many approximative tests. It is a 'gold standard' test for comparing multiple $c_v$. Our implementation is based on the following from Feltz and Miller (1996):

Let $s_i$ be the observed standard deviation of the $i$th sample (or group of measurements), $x_i$ be the observed mean of the $i$th sample, and let

$$m_i=n_i-1$$

and we do not know the population $c_v$, so let $\widehat{\tau }_c$ be an estimate as follows:

$$\widehat{\tau }_c = \frac{\left( \sum _{j=1}^km_j\frac{S_j}{\bar{x}_j} \right)}{M}$$ 
where 
$$M=\sum _{j=1}^m m_j$$ 
Then the test statistic can be computed with: 
$$\begin{aligned} D'A D= \widehat{\tau }_c^{-2}\left( 0.5+\widehat{\tau }_c^2\right) ^{-1}\left[ \sum _{i=1}^km_i\left(\frac{s_i}{\bar{x}_i} \right)^2-\frac{\widehat{\tau }_c^2}{M} \right] \end{aligned}$$
The $D'AD$ value measures how far each sample $c_v$ is from our estimate of the population $c_v$. Feltz and Miller (1996) note that the $D'AD$ value distributes as a central $\chi^2$ random variable with $k-1$ degrees of freedom, from which a p-value can be computed.

We also include the Krishnamoorthy and Lee (2014) test as a more recent development with lower rates of type I error, better performance with uneven sample numbers, and more power across a range of conditions.
